A formal treatment of bidirectional typing

### 2023-10-17 tex/ESOP24/S8-future.tex
[ tex ] Final two paragraphs of future work

\subsubsection{General theory of bidirectional type synthesis beyond simple types}

@@

\subsubsection{More characteristics of bidirectional typing}

There are still many concepts that are possible to articulate such as, for example, Pfenning's recipe for bidirectionalising typing rules.

In contrast, there are concepts that may be hard to pin down, notably `annotation character'~\cite{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs.
These considerations may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.

>>>>

\subsubsection{Beyond simple types}

@@

\subsubsection{Beyond syntax-directed systems}

For non-syntax-directed systems, since there can be multiple moding or typing rules for a language construct, both the mode decorator and the bidirectional type synthesiser will need to backtrack, and the decorator may need to produce all possible mode derivations so as not to limit the synthesiser's choices.
For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, although bidirectional typing rules and pairs of a moding rule and a typing rule may no longer in perfect correspondence, and we may need to adapt the statement of completeness (for example by taking all possible mode derivations as input to guarantee that at least one can be paired with the input typing derivation).

\subsubsection{Towards a richer formal theory}

There are more techniques in bidirectional typing that can be formally studied in general, with one notable example being the Pfenning recipe for bidirectionalising typing rules~\citep[Section~4]{Dunfield2021}.
There are also concepts that may be hard to fully formalise, for example `annotation character'~\citep[Section~3.4]{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs, but it would be interesting to explore to what extent such concepts can be formalised.

%Such aspects  may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.
